---
title: "CRC meta-analysis with MetaPhlAn 3 + HUMAnN 3"
output:
  html_document:
    df_print: paged
---
```{r include=FALSE}
library(svglite)
knitr::opts_chunk$set(echo = FALSE, fig.path = 'CRC_metaanalysis_plots/', dev = c('png','svglite'), warning = FALSE)
library(tidyverse)
library(magrittr)
library(ComplexHeatmap)
library(circlize)
library(vegan)
library(data.table)

# MTA_palette <-   c("#0039A6","#fc9665","#6CBE45","#996633","#A7A9AC","#FCCC0A","#000000","#EE352E","#00933C","#B933AD")
MTA_palette <- c("#0039A6","#fc9665","#6CBE45","#996633","#FCCC0A","#000000","#ff1919","#00933C","#B933AD")
CRC_datasets <- strsplit(read_file("CRC_datasets.lst"), '\n')[[1]]
ThomasA_2019_datasets <- c('FengQ_2015', 'YuJ_2015', 'ZellerG_2014', 'VogtmannE_2016', 'HanniganGD_2017', 'ThomasAM_2018a', 'ThomasAM_2018b')
theme_cm <- function(...){
  theme_minimal() +
  theme(text = element_text(family = "sans-serif", size = 10, colour = "black"),
        panel.grid = element_blank(),
        panel.grid.major.y = element_line(color = "gray73"),
        axis.text = element_text(family = "sans-serif", size = 10, colour = "black"),
        strip.background = element_rect(colour = "transparent", fill = "transparent"),
        strip.text = element_text(family = "sans-serif", size = 12, colour = "black")) + 
    theme(...)
  
}

```


```{r load_data, message=FALSE, warning=FALSE}
metadata <- read_tsv("tables/CRC_analysis_metadata_final_version.tsv") %>% 
            mutate(dataset_name = as.factor(dataset_name),
                   BMI = as.numeric(BMI),
                   age = as.numeric(age),
                   study_condition = factor(study_condition, labels = c('Control','CRC')),
                   gender = factor(gender),
                   country = factor(country))
metadata$dataset_name <- gsub("ThomasAM_2019_c", "YachidaS_2019", metadata$dataset_name)
metadata$dataset_name = as.factor(metadata$dataset_name)
# mpa_samples <- lapply(levels(metadata$dataset_name), function(dataset) {
#   dataset_mpa_path <- file.path('/shares/CIBIO-Storage/CM/scratch/data/meta', dataset, 'metaphlan-3.0_v30_CHOCOPhlAn_201901')
#   lapply(list.files(dataset_mpa_path), function(sample) {
#     if(sample %in% metadata$sampleID){
#       file.path(dataset_mpa_path, sample, paste0(sample,"_profile.tsv"))
#     }
#   })
# }) %>% unlist
# 
# mpa_table <- lapply(mpa_samples, function(sample_path){
#               read_tsv(sample_path, skip = 4, col_names = c('species','taxid','relative_abundance')) %>%
#                 filter(str_detect(species, 's__')) %>%
#                 mutate(species = gsub('^.*s__','',species),
#                        relative_abundance = as.double(relative_abundance),
#                        sample = gsub('_profile.tsv','',basename(sample_path))
#                      ) %>%
#                 select(-taxid)
#             }) %>%
#             bind_rows %>%
#             pivot_wider(names_from = 'sample', values_from = 'relative_abundance' , values_fill = list(relative_abundance = 0))

mpa_table <- read_tsv('tables/CRC_mpa3_statq01.tsv') %>% 
              filter(str_detect(ID, 's__')) %>% 
              mutate(ID = gsub('^.*s__','',ID))

colnames(mpa_table) <- gsub('_profile','', colnames(mpa_table))
mpa_table %<>% select(ID, metadata$sampleID)

mpa2_samples <- lapply(levels(metadata$dataset_name), function(dataset) {
  dataset_mpa_path <- file.path('/shares/CIBIO-Storage/CM/scratch/data/meta', dataset, 'metaphlan2')
  lapply(list.files(dataset_mpa_path), function(sample) {
    if(sample %in% metadata$sampleID){
      file.path(dataset_mpa_path, sample, paste0(sample,"_profile.tsv"))
    }
  })
}) %>% unlist

mpa2_table <- lapply(mpa2_samples, function(sample_path){
              read_tsv(sample_path, skip = 1, col_names = c('species','relative_abundance')) %>%
                filter(str_detect(species, 's__')) %>%
                filter(!str_detect(species, 't__')) %>%
                mutate(species = gsub('^.*s__','',species),
                       relative_abundance = as.double(relative_abundance),
                       sample = gsub('_profile.tsv','',basename(sample_path))
                     )
            }) %>%
            bind_rows %>%
            pivot_wider(names_from = 'sample', values_from = 'relative_abundance' , values_fill = list(relative_abundance = 0), values_fn = {unique})


markers <- c(
  "Streptococcus_anginosus_group",
  "Gemella_haemolysans",
  "Slackia_exigua",
  "Porphyromonas_uenonis",
  "Peptostreptococcus_anaerobius",
  "Bacteroides_nordii",
  "Selenomonas_sputigena",
  "Bacteroides_fragilis",
  "Eisenbergiella_tayi",
  "Gemella_bergeri",
  "Prevotella_intermedia",
  "Porphyromonas_asaccharolytica",
  "Bulleidia_extructa",
  "Ruthenibacterium_lactatiformans",
  "Solobacterium_moorei",
  "Dialister_pneumosintes",
  "Peptostreptococcus_stomatis",
  "Fusobacterium_nucleatum",
  "Gemella_morbillorum",
  "Parvimonas_micra"
)
```


# Table 1
```{r}
filt_metadata <- metadata %>% 
                  select(dataset_name, study_condition, country, age, gender, BMI) %>% 
                  mutate(BMI = as.numeric(BMI),
                         age = as.numeric(age),
                         study_condition = factor(study_condition),
                         gender = factor(gender),
                         country = factor(country))

filt_metadata %>% 
  group_by(dataset_name, study_condition) %>% 
  tally %>% 
  pivot_wider(names_from = 'study_condition', values_from = 'n')
```

# Alpha diversity of rarefied samples (0.1 percentile)
```{r message=FALSE, warning=FALSE}
# rarefied_mpa3_samples <- lapply(levels(metadata$dataset_name), function(dataset) {
#   dataset_mpa_path <- file.path('/shares/CIBIO-Storage/CM/scratch/data/meta', dataset, 'metaphlan-3.0_v30_CHOCOPhlAn_201901_rarefaction_10th_percentile_statq01')
#   lapply(metadata %>% filter(dataset_name == dataset) %>% pull(sampleID), function(sample) {
#     if(sample %in% metadata$sampleID){
#       path = file.path(dataset_mpa_path, sample, paste0(sample,"_rarefied_0.1_percentile_profile_statq_01.tsv"))
#       if(file.exists(path))
#         path
#     }
#   })
# }) %>% unlist
# 
# rarefied_mpa3_table <- lapply(rarefied_mpa3_samples, function(sample_path){
#               read_tsv(sample_path, skip = 4, col_names = c('species','taxid','relative_abundance')) %>%
#                 filter(str_detect(species, 's__')) %>%
#                 mutate(species = gsub('^.*s__','',species),
#                        sample = gsub('_rarefied_0.1_percentile_profile_statq_01.tsv','',basename(sample_path))
#                      ) %>%
#                 select(-taxid)
#             }) %>%
#             bind_rows %>%
#             pivot_wider(names_from = 'sample', values_from = 'relative_abundance' , values_fill = list(relative_abundance = 0))
# 
# rarefied_mpa2_samples <- lapply(levels(metadata$dataset_name), function(dataset) {
#   dataset_mpa_path <- file.path('/shares/CIBIO-Storage/CM/scratch/data/meta', dataset, 'metaphlan2_rarefacted_10percentile_reads')
#   lapply(metadata %>% filter(dataset_name == dataset) %>% pull(sampleID), function(sample) {
#     if(sample %in% metadata$sampleID){
#       path = list.files(file.path(dataset_mpa_path, sample), pattern = '.tsv', full.names = TRUE)[1]
#       if(file.exists(path))
#         path
#       }
#   })
# }) %>% unlist
# 
# rarefied_mpa2_table <- lapply(rarefied_mpa2_samples, function(sample_path){
#               read_tsv(sample_path, skip = 1, col_names = c('species','relative_abundance')) %>%
#                 filter(str_detect(species, 's__')) %>%
#                 filter(!str_detect(species, 't__')) %>%
#                 mutate(species = gsub('^.*s__','',species),
#                        sample = rev(str_split(sample_path, '/', simplify = TRUE))[2]
#                      )
#             }) %>%
#             bind_rows %>%
#             pivot_wider(names_from = 'sample', values_from = 'relative_abundance' , values_fill = list('relative_abundance' = 0))


rarefied_mpa3_table <- read_tsv('tables/CRC_mpa3_rarefied_0.1_percentile_statq_01.tsv') %>% 
  filter(str_detect(ID, 's__')) %>% 
  mutate(ID = gsub('^.*s__','',ID))
colnames(rarefied_mpa3_table) <- gsub('_rarefied_0.1_percentile_profile_statq_01','',colnames(rarefied_mpa3_table))

rarefied_mpa3_table_statq_02 <- read_tsv('tables/CRC_mpa3_rarefied_0.1_percentile_statq_02.tsv') %>% 
  filter(str_detect(ID, 's__')) %>% 
  mutate(ID = gsub('^.*s__','',ID))
colnames(rarefied_mpa3_table_statq_02) <- gsub('_rarefied_0.1_percentile_profile','',colnames(rarefied_mpa3_table_statq_02))

rarefied_mpa2_table <- read_tsv('tables/CRC_mpa2_rarefied_0.1_percentile.tsv') %>% 
                          filter(str_detect(ID, 's__')) %>% 
                          filter(!str_detect(ID, 't__')) %>% 
                          mutate(ID = gsub('^.*s__','',ID))
colnames(rarefied_mpa2_table) <- gsub('_metaphlan2.7_rarefied_0.1_percentile_profile','',colnames(rarefied_mpa2_table))
colnames(rarefied_mpa2_table) <- gsub('_profile','',colnames(rarefied_mpa2_table))

rarefied_mpa2_table_statq_02 <- read_tsv('tables/CRC_mpa2_rarefied_0.1_percentile_statq_02.tsv')

shared_samples <- inner_join(metadata %>% filter(sampleID %in% colnames(rarefied_mpa2_table)[-1]), metadata %>% filter(sampleID %in% colnames(rarefied_mpa3_table)[-1]))$sampleID

rarefied_mpa2_table %<>% select(ID, shared_samples)
rarefied_mpa3_table %<>% select(ID, shared_samples)

shared_samples <- inner_join(metadata %>% filter(sampleID %in% colnames(rarefied_mpa2_table_statq_02)[-1]), metadata %>% filter(sampleID %in% colnames(rarefied_mpa3_table_statq_02)[-1]))$sampleID

rarefied_mpa2_table_statq_02 %<>% select(species, shared_samples)
rarefied_mpa3_table_statq_02 %<>% select(ID, shared_samples)

nspecies_mpa3 <- tibble(sample = colnames(rarefied_mpa3_table[,-1]), count = colSums(rarefied_mpa3_table[,-1] > 0 ))
nspecies_mpa2 <- tibble(sample = colnames(rarefied_mpa2_table[,-1]), count = colSums(rarefied_mpa2_table[,-1] > 0 ))
  
nspecies_mpa3_statq_02 <- tibble(sample = colnames(rarefied_mpa3_table_statq_02[,-1]), count = colSums(rarefied_mpa3_table_statq_02[,-1] > 0 ))
nspecies_mpa2_statq_02 <- tibble(sample = colnames(rarefied_mpa2_table_statq_02[,-1]), count = colSums(rarefied_mpa2_table_statq_02[,-1] > 0 ))
```

```{r}
left_join(nspecies_mpa3, metadata, by = c('sample' = 'sampleID')) %>%
select(sample, count, dataset_name, study_condition) %>%
group_by(dataset_name) %>%
summarize(wilcox = wilcox.test(count ~ study_condition)$p.value %>% signif(2))
```

# Median percent increase compared to MetaPhlAn2 using stat_q 0.1
```{r}
median_0.1 <- full_join(nspecies_mpa2, nspecies_mpa3, suffix = c("_mpa2","_mpa3"), by = "sample") %>%
  left_join(metadata, by = c("sample" = "sampleID")) %>%
  select(contains("count_"), sample, dataset_name, study_condition) %>%
  group_by(dataset_name, study_condition) %>%
  summarize(median_mpa2 = median(count_mpa2), median_mpa3 = median(count_mpa3)) %>%
  mutate(perc_inc = ((median_mpa3-median_mpa2)/median_mpa2 * 100) %>% round(1))
median_0.1
```
  
# Median percent increase compared to MetaPhlAn2 using stat_q 0.2
```{r}
median_0.2 <- full_join(nspecies_mpa2_statq_02, nspecies_mpa3_statq_02, suffix = c("_mpa2","_mpa3"), by = "sample") %>%
  left_join(metadata, by = c("sample" = "sampleID")) %>%
  select(contains("count_"), sample, dataset_name, study_condition) %>%
  group_by(dataset_name, study_condition) %>%
  summarize(median_mpa2 = median(count_mpa2), median_mpa3 = median(count_mpa3)) %>%
  mutate(perc_inc = ((median_mpa3-median_mpa2)/median_mpa2 * 100) %>% round(1))
median_0.2
```

```{r}
write_tsv(median_0.1, 'tables/mpa3_mpa2_statq_0.1_alphadiv_median_increase.tsv')
write_tsv(median_0.2, 'tables/mpa3_mpa2_statq_0.2_alphadiv_median_increase.tsv')
```


```{r metaphlan3_alpha_diversity_statq01, fig.height=3}
left_join(nspecies_mpa3, metadata, by = c('sample' = 'sampleID')) %>% 
  select(sample, count, dataset_name, study_condition) %>% 
  ggplot(aes(reorder(dataset_name, -count, mean) ,count)) +
  stat_boxplot(aes(fill=study_condition)) + 
  theme_cm(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1.05), axis.line = element_line()) + 
  scale_fill_manual(values = c('#758862', '#AC5248'), name = 'Condition') + 
  xlab('Dataset name') +
  ylab('Number of species')
```

```{r}
mpa_table_markers <-
  mpa_table %>%
  filter(ID %in% markers) %>%
  arrange(desc(rowMeans(.[,-1]))) %>% 
  select(ID, noquote(metadata %>% arrange(study_condition) %>% select(sampleID)) %>% t %>% as.character) %>% 
  mutate_if(is.numeric, ~ log10(1+.)) %>% 
  mutate(ID = gsub('_',' ', ID)) %>% 
  column_to_rownames('ID')
```


# MDS
```{r calculate MDS}
library(MMUPHin)
mat_sp <- mpa_table[, -1] %>% 
  as.matrix() %>% 
  apply(2, function(x) x / sum(x))
rownames(mat_sp) <- mpa_table$ID

df_meta <- metadata %>% 
  as.data.frame() %>% 
  tibble::column_to_rownames("sampleID")

mat_sp <- mat_sp[, rownames(df_meta)]

fit_adjust_batch <- MMUPHin::adjust_batch(mat_sp,
                                          batch = "dataset_name",
                                          covariates = "study_condition",
                                          data = df_meta)
mat_sp_norm <- fit_adjust_batch$feature_abd_adj

mpa_tree <- ape::read.tree('../tree_taxa_reroot.nwk')
mpa_tree$tip.label <- gsub(".+\\|s__", "", mpa_tree$tip.label)

filt_tree <- ape::keep.tip(mpa_tree, intersect(rownames(mat_sp_norm),mpa_tree$tip.label))
filt_mpa_table <- mat_sp_norm[filt_tree$tip.label,]

rbiom_weighted <- rbiom::unifrac(asinh(sqrt(filt_mpa_table)), weighted=TRUE, tree=filt_tree)

mds_res_nonorm <- cmdscale(rbiom_weighted)
```

```{r CRC_MDS_no_normalization, fig.width=8, fig.height=4, fig.retina=TRUE}
mds_res_nonorm %>% 
  as.data.frame() %>% 
  rownames_to_column('sampleID') %>% 
  left_join(metadata, by = 'sampleID') %>% 
  select(sampleID, MDS1=V1, MDS2=V2, study_condition, dataset_name) %>% 
ggplot() +
  geom_point(aes(MDS1,MDS2, shape=study_condition, color=dataset_name), size=1.5) +
  scale_color_manual(values = MTA_palette, name = 'Dataset') +
  theme_minimal(base_size = 15) + 
  scale_shape_discrete(name = 'Condition')
```

# Heatmap: Abundance of CRC Biomarkers
```{r}
annot_v <- metadata %>% 
  arrange(study_condition) %>%  
  select(sampleID, Condition=study_condition, Dataset=dataset_name) %>% 
  mutate(Condition =  sub("(.)", "\\U\\1", Condition, perl=TRUE)) %>% 
  column_to_rownames("sampleID")

cols <- list("Condition" = c("Control" = "#758862", "CRC" = "#AC5248"),
             "Dataset" = setNames(MTA_palette, sort(unique(annot_v$Dataset)))
            )
```

```{r metaphlan3_CRC_biomarkers, fig.width=10, fig.height=5}
x <- colorRamp2(breaks = seq(0,1,0.1), colors = hcl.colors(11, palette = 'Inferno'), transparency = 0.1, )
attributes(x)$breaks <- seq(0,2,0.1)
Heatmap(as.matrix(mpa_table_markers),
        top_annotation = HeatmapAnnotation(df = annot_v, col = cols),
        col = x,
        name = "Relative abundance(Log))",
        column_split = annot_v$Condition,
        column_gap = unit(1,"points"),
        column_title = rep("",2),
        cluster_rows = FALSE,
        show_column_names = FALSE,
        cluster_column_slices = TRUE,
        show_column_dend = FALSE,
        row_names_gp = gpar(fontface = 'italic', fontfamily = "sans"),
        heatmap_legend_param = list(direction = 'horizontal', border = 'black', position = 'top')
)

```

# Heatmap: Relative abundance of top 10 species
```{r}
top30_mpa_table <- mpa_table %>% 
  select(ID, noquote(metadata %>% arrange(study_condition) %>% select(sampleID)) %>% t %>% as.character) %>% 
  # mutate_if(is.numeric, funs(log10(1+.))) %>% 
  mutate(ID = gsub('_',' ', ID)) %>% 
  top_n(10, wt = rowMeans(.[,-1])) %>% 
  column_to_rownames('ID')


annot_v <- metadata %>% 
  arrange(study_condition) %>%  
  select(sampleID, Condition=study_condition, Dataset=dataset_name) %>% 
  mutate(Condition =  sub("(.)", "\\U\\1", Condition, perl=TRUE)) %>% 
  column_to_rownames("sampleID")

cols <- list("Condition" = c("Control" = "#758862", "CRC" = "#AC5248"),
             "Dataset" = setNames(MTA_palette, sort(unique(annot_v$Dataset)))
            )
```

```{r metaphlan3_CRC_top30, fig.width=10, fig.height=5}
x <- colorRamp2(breaks = seq(0,2,0.1), colors = hcl.colors(21, palette = 'Inferno'), transparency = 0.1, )
ComplexHeatmap::Heatmap(log10(1+as.matrix(top30_mpa_table)), 
        top_annotation = HeatmapAnnotation(df = annot_v, col = cols),
        col = x,
        name = "Relative abundance(Log))",
        column_split = annot_v$Condition,
        column_gap = unit(1,"points"),
        column_title = rep("",2),
        cluster_rows = FALSE,
        show_column_names = FALSE,
        cluster_column_slices = TRUE,
        show_column_dend = FALSE,
        row_names_gp = gpar(fontface = 'italic', fontfamily = "sans"),
        heatmap_legend_param = list(direction = 'horizontal', border = 'black', position = 'top')
)
```

## HUMAnN 3

### Normalize per CPM
```{bash eval=FALSE, include=FALSE}
conda activate humann-3.0

for dataset in FengQ_2015 GuptaA_2019 ThomasAM_2019_a ThomasAM_2019_b ThomasAM_2019_c VogtmannE_2016 WirbelJ_2018 YachidaS_2019 YuJ_2015 ZellerG_2014
do
  find /shares/CIBIO-Storage/CM/scratch/data/meta/${dataset}/humann-3.0_v30_CHOCOPhlAn_201901 -name *_pathabundance.tsv | xargs -I {} -P 100 bash -c 'x={}; humann_renorm_table --input ${x} --units cpm --update-snames --output ${x/pathabundance/pathabundance-cpm} 2>/dev/null'
done

for dataset in FengQ_2015 GuptaA_2019 ThomasAM_2019_a ThomasAM_2019_b ThomasAM_2019_c VogtmannE_2016 WirbelJ_2018 YachidaS_2019 YuJ_2015 ZellerG_2014
do
  find /shares/CIBIO-Storage/CM/scratch/data/meta/${dataset}/humann-3.0_v30_CHOCOPhlAn_201901 -name *_genefamilies.tsv | xargs -I {} -P 100 bash -c 'x={}; humann_renorm_table --input ${x} --units cpm --update-snames --output ${x/genefamilies/genefamilies-cpm} 2>/dev/null'
done

for dataset in FengQ_2015 GuptaA_2019 ThomasAM_2019_a ThomasAM_2019_b ThomasAM_2019_c VogtmannE_2016 WirbelJ_2018 YachidaS_2019 YuJ_2015 ZellerG_2014
do
  find /shares/CIBIO-Storage/CM/scratch/data/meta/${dataset}/humann-3.0_v30_CHOCOPhlAn_201901 -name *_genefamilies.tsv | xargs -I {} -P 100 bash -c 'x={}; humann_regroup_table --input ${x} --output ${x/genefamilies/genefamilies-ecgrouped} --group uniref90_level4ec -e 10'
done

for dataset in FengQ_2015 GuptaA_2019 ThomasAM_2019_a ThomasAM_2019_b ThomasAM_2019_c VogtmannE_2016 WirbelJ_2018 YachidaS_2019 YuJ_2015 ZellerG_2014
do
  find /shares/CIBIO-Storage/CM/scratch/data/meta/${dataset}/humann-3.0_v30_CHOCOPhlAn_201901 -name *_genefamilies-ecgrouped.tsv | xargs -I {} -P 100 bash -c 'x={}; humann_renorm_table --input ${x} --units cpm --update-snames --output ${x/genefamilies-ecgrouped/genefamilies-ecgrouped-cpm_norm}'
done
```


```{r load_humann_pwy, message=FALSE, warning=FALSE}
CRC_datasets <- c("FengQ_2015", "GuptaA_2019", "ThomasAM_2019_a", "ThomasAM_2019_b", "VogtmannE_2016", "WirbelJ_2018", "YachidaS_2019", "YuJ_2015", "ZellerG_2014")

# hnn_pathabundance <- lapply(CRC_datasets, function(dataset) { 
#   dataset_hnn_path <- file.path('/shares/CIBIO-Storage/CM/scratch/data/meta', dataset, 'humann-3.0_v30_CHOCOPhlAn_201901')
#   lapply(list.files(dataset_hnn_path), function(sample) {
#     if(sample %in% metadata$sampleID){
#       file.path(dataset_hnn_path, sample, paste0(sample,"_pathabundance-cpm.tsv"))
#     }
#   })
# }) %>% unlist
# 
# hnn_normalizedgenefamilies <- lapply(CRC_datasets, function(dataset) {
#   dataset_hnn_path <- file.path('/shares/CIBIO-Storage/CM/scratch/data/meta', dataset, 'humann-3.0_v30_CHOCOPhlAn_201901')
#   lapply(list.files(dataset_hnn_path), function(sample) {
#     if(sample %in% metadata$sampleID){
#       path=file.path(dataset_hnn_path, sample, paste0(sample,"_genefamilies-cpm.tsv"))
#       if(file.exists(path))
#         path
#     }
#   })
# }) %>% unlist
# 
# hnn_normalizedEC <- lapply(CRC_datasets, function(dataset) {
#   dataset_hnn_path <- file.path('/shares/CIBIO-Storage/CM/scratch/data/meta', dataset, 'humann-3.0_v30_CHOCOPhlAn_201901')
#   lapply(list.files(dataset_hnn_path), function(sample) {
#     if(sample %in% metadata$sampleID){
#       file.path(dataset_hnn_path, sample, paste0(sample,"_genefamilies-ecgrouped-cpm_norm.tsv"))
#     }
#   })
# }) %>% unlist
# 
# hnn_pathcoverage <- lapply(CRC_datasets, function(dataset) {
#   dataset_hnn_path <- file.path('/shares/CIBIO-Storage/CM/scratch/data/meta', dataset, 'humann-3.0_v30_CHOCOPhlAn_201901')
#   lapply(list.files(dataset_hnn_path), function(sample) {
#     if(sample %in% metadata$sampleID){
#     file.path(dataset_hnn_path, sample, paste0(sample,"_pathcoverage.tsv"))
#     }
#   })
# }) %>% unlist

cutC_NR90_ids <- strsplit(read_file("cutC_NR90"), '\n')[[1]]
cutD_NR90_ids <- strsplit(read_file("cutD_NR90"), '\n')[[1]]
yeaW_NR90_ids <- strsplit(read_file("yeaW_NR90"), '\n')[[1]]

# hnn_pwyab <- data.table::rbindlist(mclapply(hnn_pathabundance, function(sample_path) {
#                             data.table::fread(cmd = paste0('grep -v "|" ', sample_path), nThread = 5, col.names = c('Pathway','relative_abundance')) %>% 
#                               cbind(sample = gsub("_pathabundance.tsv","",basename(sample_path)))
#                   }, mc.cores = 30))
# hnn_pwyab <- dcast(hnn_pwyab, Pathway ~ sample, value.var = "relative_abundance", fill = 0)

# hnn_genefamilies <- data.table::rbindlist(mclapply(hnn_normalizedgenefamilies, function(sample_path) {
#                             data.table::fread(cmd = paste0('grep -v "|" ', sample_path), nThread = 5, col.names = c('GeneFamily','relative_abundance')) %>% 
#                               cbind(sample = gsub("_genefamilies-cpm.tsv","",basename(sample_path)))
#                   }, mc.cores = 30))
# hnn_genefamilies <- dcast(hnn_genefamilies, GeneFamily ~ sample, value.var = "relative_abundance", fill = 0)

# hnn_ec <- data.table::rbindlist(mclapply(hnn_normalizedEC, function(sample_path) {
#                             data.table::fread(cmd = paste0('grep -v "|" ', sample_path), nThread = 5, col.names = c('EC','relative_abundance')) %>% 
#                               cbind(sample = gsub("_genefamilies-ecgrouped-cpm_norm.tsv","",basename(sample_path)))
#                   }, mc.cores = 30))
# hnn_ec <- dcast(hnn_ec, EC ~ sample, value.var = "relative_abundance", fill = 0)

# hnn_pwycov <- lapply(hnn_pathcoverage, function(sample_path) {
#                           if (file.exists(sample_path)){
#                             read_tsv(sample_path) %>%
#                               pivot_longer(-1, names_to = 'sample', values_to = 'coverage') %>%
#                               mutate(sample = str_replace(sample, '_Coverage',''))
#                           }
#                   }) %>%
#               bind_rows %>%
#               pivot_wider(names_from = 'sample', values_from = 'coverage')

hnn_pwyab <- data.table::fread('tables/humann3_pathabundance_CRC_cohorts_cpm_unstratified.tsv')
colnames(hnn_pwyab) <- gsub('_Abundance','', colnames(hnn_pwyab))
hnn_pwyab %<>% select(Pathway, metadata$sampleID)

hnn_ec <- data.table::fread('tables/CRC_humann3_ecs_cpm_ungrouped.tsv')
colnames(hnn_ec) <- gsub('_Abundance-CPM','', colnames(hnn_ec))
colnames(hnn_ec)[1] <- 'EC'
hnn_ec %<>% select(EC, metadata$sampleID)
```


# MDS: Pathways
```{r}
mds_pwy <- cmdscale(vegdist(t(hnn_pwyab[,-1]), diag = TRUE, upper = FALSE, method = 'bray'))
```


```{r CRC_MDS_pwys, fig.width=8, fig.height=4.5, fig.retina=TRUE}
mds_pwy %>% 
  as.data.frame() %>% 
  rownames_to_column('sampleID') %>% 
  left_join(metadata, by = 'sampleID') %>% 
  select(sampleID, MDS1=V1, MDS2=V2, study_condition, dataset_name) %>% 
ggplot() +
  geom_point(aes(MDS1,MDS2, shape=study_condition, color=dataset_name), size=1.5) +
  scale_color_manual(values = MTA_palette, name = 'Dataset') +
  theme_minimal(base_size = 15) + 
  scale_shape_discrete(name = 'Condition')
```

# MDS: ECs
```{r}
mds_ecs <- cmdscale(vegdist(t(hnn_ec[,-1]), diag = TRUE, upper = FALSE, method = 'bray'), )
```


```{r CRC_MDS_ecs, fig.width=8, fig.height=4.5, fig.retina=TRUE}
mds_ecs %>% 
  as.data.frame() %>% 
  rownames_to_column('sampleID') %>% 
  left_join(metadata, by = 'sampleID') %>% 
  select(sampleID, MDS1=V1, MDS2=V2, study_condition, dataset_name) %>% 
ggplot() +
  geom_point(aes(MDS1,MDS2, shape=study_condition, color=dataset_name), size=1.5) +
  scale_color_manual(values = MTA_palette, name = 'Dataset') +
  theme_minimal(base_size = 15) + 
  scale_shape_discrete(name = 'Condition')
```


```{bash eval=FALSE, include=FALSE}
for dataset in FengQ_2015 GuptaA_2019 ThomasAM_2019_a ThomasAM_2019_b VogtmannE_2016 WirbelJ_2018 YachidaS_2019 YuJ_2015 ZellerG_2014
do
  find /shares/CIBIO-Storage/CM/scratch/data/meta/${dataset}/humann-3.0_v30_CHOCOPhlAn_201901 -name *_genefamilies-cpm.tsv | xargs -I {} -P 100 bash -c 'x={}; grep -Fwf /shares/CIBIO-Storage/CM/news/users/francesco.beghini/chocophlan_paper/CRC_metaphlan3/cutC_NR90 $x | grep -v "|" | awk -v a=${x##*/} "{print \$0 \"\t\" a }"'
done > merged_cutC_RPK_CPM.tsv

for dataset in FengQ_2015 GuptaA_2019 ThomasAM_2019_a ThomasAM_2019_b VogtmannE_2016 WirbelJ_2018 YachidaS_2019 YuJ_2015 ZellerG_2014
do
  find /shares/CIBIO-Storage/CM/scratch/data/meta/${dataset}/humann-3.0_v30_CHOCOPhlAn_201901 -name *_genefamilies-cpm.tsv | xargs -I {} -P 100 bash -c 'x={}; grep -Fwf /shares/CIBIO-Storage/CM/news/users/francesco.beghini/chocophlan_paper/CRC_metaphlan3/cutD_NR90 $x | grep -v "|" | awk -v a=${x##*/} "{print \$0 \"\t\" a }"'
done > merged_cutD_RPK_CPM.tsv

for dataset in FengQ_2015 GuptaA_2019 ThomasAM_2019_a ThomasAM_2019_b VogtmannE_2016 WirbelJ_2018 YachidaS_2019 YuJ_2015 ZellerG_2014
do
  find /shares/CIBIO-Storage/CM/scratch/data/meta/${dataset}/humann-3.0_v30_CHOCOPhlAn_201901 -name "*_genefamilies-cpm.tsv" | xargs -I {} -P 100 bash -c 'x={}; grep -Fwf /shares/CIBIO-Storage/CM/news/users/francesco.beghini/chocophlan_paper/CRC_metaphlan3/yeaW_NR90 $x | grep -v "|" | awk -v a=${x##*/} "{print \$0 \"\t\" a }"'
done > merged_yeaW_RPK_CPM.tsv
```


```{r load_cutC_RPK, message=FALSE, warning=FALSE}
NR90_cutC <- 
  read_tsv("merged_cutC_RPK_CPM.tsv", col_names = c('NR90', 'CPM', 'sample')) %>% 
  filter(NR90 %in% cutC_NR90_ids) %>% 
  mutate(sample = str_remove(sample, "_genefamilies.*")) %>% 
  filter(sample %in% metadata$sampleID) %>% 
  group_by(sample) %>% 
  summarise(CPM = sum(CPM)) %>% 
  bind_rows(tibble(sample=metadata$sampleID[!metadata$sampleID %in% .$sample],CPM=0))
```

```{r load_cutD_RPK, message=FALSE, warning=FALSE}
NR90_cutD <- read_tsv("merged_cutD_RPK_CPM.tsv", col_names = c('NR90', 'CPM', 'sample')) %>% 
  filter(NR90 %in% cutD_NR90_ids) %>% 
  mutate(sample = str_remove(sample, "_genefamilies.*")) %>% 
  filter(sample %in% metadata$sampleID) %>%
  group_by(sample) %>% 
  summarise(CPM = sum(CPM)) %>% 
  bind_rows(tibble(sample=metadata$sampleID[!metadata$sampleID %in% .$sample],CPM=0))
```

```{r load_yeaW_RPK, message=FALSE, warning=FALSE}
NR90_yeaW <- read_tsv("merged_yeaW_RPK_CPM.tsv", col_names = c('NR90', 'CPM', 'sample')) %>% 
  filter(NR90 %in% yeaW_NR90_ids) %>% 
  mutate(sample = str_remove(sample, "_genefamilies.*")) %>% 
  filter(sample %in% metadata$sampleID) %>% 
  group_by(sample) %>% 
  summarise(CPM = sum(CPM)) %>% 
  bind_rows(tibble(sample=metadata$sampleID[!metadata$sampleID %in% .$sample],CPM=0))
```

## cutC UniRef90 Wilcoxon per dataset CRC vs Controls
```{r}
NR90_cutC %>%
  full_join(metadata, by = c('sample' = 'sampleID')) %>% 
  group_by(dataset_name) %>%
  summarise(wilcox = wilcox.test(CPM ~ study_condition, paired=FALSE)$p.value %>% round(4))
```


## cutC EC 4.3.99.4 Wilcoxon per dataset CRC vs Controls
```{r}
hnn_ec %>%
  filter(EC == '4.3.99.4') %>%
  pivot_longer(-EC, names_to = 'sample', values_to = 'CPM') %>%
  select(-EC) %>% 
  full_join(metadata, by = c('sample' = 'sampleID')) %>% 
  group_by(dataset_name) %>%
  summarise(wilcox = wilcox.test(CPM ~ study_condition, paired=FALSE)$p.value %>% round(4))
```

## cutD UniRef90 Wilcoxon per dataset CRC vs Controls
```{r}
NR90_cutD %>%
  full_join(metadata, by = c('sample' = 'sampleID')) %>% 
  group_by(dataset_name) %>%
  summarise(wilcox = wilcox.test(CPM ~ study_condition, paired=FALSE)$p.value %>% round(4))
```

## yeaW UniRef90 Wilcoxon per dataset CRC vs Controls
```{r}
NR90_yeaW %>%
  full_join(metadata, by = c('sample' = 'sampleID')) %>% 
  group_by(dataset_name) %>%
  summarise(wilcox = wilcox.test(CPM ~ study_condition, paired=FALSE)$p.value %>% round(4))
```

## Distribution of cutC CPM per dataset
```{r NR90_cutC_CPM, fig.height=3}
NR90_cutC %>%
  full_join(metadata, by = c('sample' = 'sampleID')) %>%
  ggplot(aes(dataset_name,log(1+CPM))) +
  stat_boxplot(aes(fill=study_condition)) + 
  theme_cm(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1.05), axis.line = element_line()) + 
  scale_fill_manual(values = c('#758862', '#AC5248'), name = 'Condition') + 
  xlab('Dataset name')
```

## Distribution of yeaW CPM per dataset
```{r NR90_yeaW_CPM, fig.height=3, fig.width=6}
NR90_yeaW %>%
  full_join(metadata, by = c('sample' = 'sampleID')) %>%
  ggplot(aes(dataset_name,log(1+CPM))) +
  stat_boxplot(aes(fill=study_condition)) + 
  theme_cm(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1.05), axis.line = element_line()) + 
  scale_fill_manual(values = c('#758862', '#AC5248'), name = 'Condition') + 
  xlab('Dataset name')
```

```{r eval=FALSE, fig.height=4, fig.width=9, include=FALSE}
ggplot(effect_sizes) + 
  geom_point(size = 3.5, aes(Species, value, shape = variable, color= variable)) +
  geom_linerange(aes(Species, value,ymin=CI_Low, ymax=CI_UP)) + 
  scale_shape_manual(name = 'Dataset', values= c(19, 8, 11, 5, 2, 0, 13, 12, 4, 3)) + 
  scale_color_manual(name = 'Dataset', values = c('black', MTA_palette)) +
  coord_flip() +
  theme_minimal() +
  geom_hline(aes(yintercept = 0), color = "gray", linetype = "dashed") +
  theme(axis.title.y = element_blank(), 
        axis.title.x = element_text(family = "sans-serif", size = 7, colour = "black"),
        axis.text = element_text(family = "sans-serif", size = 7, colour = "black"),
        strip.text = element_text(family = "sans-serif", size = 12, colour = "black")
        ) +
  guides(shape = FALSE, color = FALSE)+
  ylab("Effect size") 
```


## ML heatmap
```{r ml_heatmap, fig.width=7}
ML <- tribble(~source, ~dataset, ~AUC, ~type, 
              "MetaPhlAn 3\nspecies", "ZellerG_2014", .82, "LODO",
              "MetaPhlAn 3\nspecies", "YuJ_2015", .90, "LODO",
              "MetaPhlAn 3\nspecies", "FengQ_2015", .88, "LODO",
              "MetaPhlAn 3\nspecies", "VogtmannE_2016", 0.82, "LODO",
              "MetaPhlAn 3\nspecies", "WirbelJ_2018", 0.91, "LODO",
              "MetaPhlAn 3\nspecies", "GuptaA_2019", 0.97, "LODO",
              "MetaPhlAn 3\nspecies", "ThomasAM_2019_a", 0.76, "LODO",
              "MetaPhlAn 3\nspecies", "ThomasAM_2019_b", 0.85, "LODO",
              "MetaPhlAn 3\nspecies", "YachidaS_2019", 0.76, "LODO",
              
              "HUMAnN 3\nE.C.", "ZellerG_2014", 0.7646, "LODO",
              "HUMAnN 3\nE.C.", "YuJ_2015", 0.7147, "LODO",
              "HUMAnN 3\nE.C.", "FengQ_2015", 0.732, "LODO",
              "HUMAnN 3\nE.C.", "VogtmannE_2016", 0.6405, "LODO",
              "HUMAnN 3\nE.C.", "WirbelJ_2018", 0.7385, "LODO",
              "HUMAnN 3\nE.C.", "GuptaA_2019", 0.904, "LODO",
              "HUMAnN 3\nE.C.", "ThomasAM_2019_a", 0.8813, "LODO",
              "HUMAnN 3\nE.C.", "ThomasAM_2019_b", 0.6472, "LODO",
              "HUMAnN 3\nE.C.", "YachidaS_2019", 0.6442, "LODO",
              
              "HUMAnN 3\nGene Family","ZellerG_2014", 0.8295, "LODO",
              "HUMAnN 3\nGene Family","YuJ_2015", 0.8158, "LODO",
              "HUMAnN 3\nGene Family","FengQ_2015", 0.8386, "LODO",
              "HUMAnN 3\nGene Family","VogtmannE_2016", 0.7097, "LODO",
              "HUMAnN 3\nGene Family","WirbelJ_2018", 0.8060, "LODO",
              "HUMAnN 3\nGene Family","GuptaA_2019", 0.9359, "LODO",
              "HUMAnN 3\nGene Family","ThomasAM_2019_a", 0.8250, "LODO",
              "HUMAnN 3\nGene Family","ThomasAM_2019_b", 0.7275, "LODO",
              "HUMAnN 3\nGene Family","YachidaS_2019", 0.6871, "LODO",
              
              "HUMAnN 3\nPathways", "ZellerG_2014",0.7655, "LODO",
              "HUMAnN 3\nPathways", "YuJ_2015",0.6815, "LODO",
              "HUMAnN 3\nPathways", "FengQ_2015",0.7207, "LODO",
              "HUMAnN 3\nPathways", "VogtmannE_2016",0.6272, "LODO",
              "HUMAnN 3\nPathways", "WirbelJ_2018",0.7398, "LODO",
              "HUMAnN 3\nPathways", "GuptaA_2019",0.8793, "LODO",
              "HUMAnN 3\nPathways", "ThomasAM_2019_a",0.8122, "LODO",
              "HUMAnN 3\nPathways", "ThomasAM_2019_b",0.6993, "LODO",
              "HUMAnN 3\nPathways", "YachidaS_2019",0.6089, "LODO",
              
              
              "MetaPhlAn 3\nspecies", "ZellerG_2014", .87, "CV",
              "MetaPhlAn 3\nspecies", "YuJ_2015", .89, "CV",
              "MetaPhlAn 3\nspecies", "FengQ_2015", .90, "CV",
              "MetaPhlAn 3\nspecies", "VogtmannE_2016", 0.68, "CV",
              "MetaPhlAn 3\nspecies", "WirbelJ_2018", 0.93, "CV",
              "MetaPhlAn 3\nspecies", "GuptaA_2019", 0.96, "CV",
              "MetaPhlAn 3\nspecies", "ThomasAM_2019_a", 0.74, "CV",
              "MetaPhlAn 3\nspecies", "ThomasAM_2019_b", 0.78, "CV",
              "MetaPhlAn 3\nspecies", "YachidaS_2019", 0.85, "CV",
              
              "HUMAnN 3\nE.C.", "ZellerG_2014", 0.83, "CV",
              "HUMAnN 3\nE.C.", "YuJ_2015", 0.76, "CV",
              "HUMAnN 3\nE.C.", "FengQ_2015", 0.79, "CV",
              "HUMAnN 3\nE.C.", "VogtmannE_2016", 0.61, "CV",
              "HUMAnN 3\nE.C.", "WirbelJ_2018", 0.78, "CV",
              "HUMAnN 3\nE.C.", "GuptaA_2019", 0.96, "CV",
              "HUMAnN 3\nE.C.", "ThomasAM_2019_a", 0.75, "CV",
              "HUMAnN 3\nE.C.", "ThomasAM_2019_b", 0.73, "CV",
              "HUMAnN 3\nE.C.", "YachidaS_2019", 0.79, "CV",
              
              "HUMAnN 3\nGene Family","ZellerG_2014", 0.7949, "CV",
              "HUMAnN 3\nGene Family","YuJ_2015", 0.81, "CV",
              "HUMAnN 3\nGene Family","FengQ_2015", 0.78, "CV",
              "HUMAnN 3\nGene Family","VogtmannE_2016", 0.63, "CV",
              "HUMAnN 3\nGene Family","WirbelJ_2018", 0.92, "CV",
              "HUMAnN 3\nGene Family","GuptaA_2019", 0.9522, "CV",
              "HUMAnN 3\nGene Family","ThomasAM_2019_a", .6908, "CV",
              "HUMAnN 3\nGene Family","ThomasAM_2019_b", 0.71, "CV",
              "HUMAnN 3\nGene Family","YachidaS_2019", 0.8234, "CV",
              
              "HUMAnN 3\nPathways", "ZellerG_2014",0.8121, "CV",
              "HUMAnN 3\nPathways", "YuJ_2015",0.6698, "CV",
              "HUMAnN 3\nPathways", "FengQ_2015",0.7106, "CV",
              "HUMAnN 3\nPathways", "VogtmannE_2016",0.6259, "CV",
              "HUMAnN 3\nPathways", "WirbelJ_2018",0.7569, "CV",
              "HUMAnN 3\nPathways", "GuptaA_2019",0.9406, "CV",
              "HUMAnN 3\nPathways", "ThomasAM_2019_a",0.7583, "CV",
              "HUMAnN 3\nPathways", "ThomasAM_2019_b",0.7147, "CV",
              "HUMAnN 3\nPathways", "YachidaS_2019",0.7554, "CV",

    )
ML$dataset <- factor(ML$dataset)
ML %>% 
ggplot(aes(y = type, x=dataset)) + 
  geom_tile(aes(fill=AUC)) + 
  geom_text(aes(label=signif(AUC,2), color= AUC < 0.73), size=3.7 ) +
  theme_cm(panel.grid.major.y = element_blank(),
           strip.text.y.right = element_text(family = "sans-serif", size = 12, colour = "black", angle = 0),
           legend.text = element_text(size = 10),
           axis.text.x = element_text(size = 12, angle = 45, vjust = 1, hjust = 1),
           legend.title = element_text(size = 10),
           panel.spacing.x = unit(0.1,'lines')
         ) +
  scale_fill_gradientn(values = c(0, 0.3, 0.6, 1), colors = c('#000000', '#E60000', '#FFD200', '#FFFFFF'), limits = c(0.5, 1))+
  scale_color_manual(guide = FALSE, values = c("black", "white"))+
  facet_grid(fct_rev(source) ~ ., space = 'free') +
  xlab('') + ylab('')
```
